Network based multifactorial modelling microRNA-target interaction.

### 2019-09-15 article.Rmd
last meeting

---
title: "Network based multifactorial modelling microRNA-target interaction."
author:
- Selcen Ari^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
- Alper Yilmaz^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
  html_document: default
documentclass: article
link-citations: yes
cls: NAR.cls
template: NAR-sample.tex
bibliography: ref.bib
---

@@

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs. We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *ABCC1*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package [ceRNAnetsim](https://github.com/selcenari/ceRNAnetsim).

@@

### Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (\autoref{fig1}a) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (\autoref{fig1}b). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (\autoref{fig1}d), consequently have potential to effect whole network due to "ripple effect".

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). c) Distribution of miRNA1 (M1) changes. d) The change at expression of common target effects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.](Fig1.jpg){#fig1}

@@

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficieny of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically effect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network: https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (\autoref{fig1}d).

@@

[#TODO]: <> "figure1, 2 ve 3 uzerinden farkli dagilim ve degradasyon açıklaması buraya. G2 uzerinden olmasa iyi olur"

@@

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary File(fig1_2app.Rmd)**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

@@

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. Simulations show that change in expression level of single gene has potential to effect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

[#TODO] We have observed that the results of simulation with ABCC1 gene, *Multidrug resistance-associated protein 1*, which is one of the most significant factors to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach. Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

[#TODO]: <> "run perturbation screening on tcga set, find a gene then couple sentences about that gene"

@@

Network based approaches for analyzing miRNA:target interactions have been developed in earlier studies. An initial  attempt demonstrated the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs [@figliuzzi_micrornas_2013]. Next, a network based kinetic model integrating miRNA and target rates of transcription, degradation, binding and unbinding was developed [@nitzan_interactions_2014] using high throughput experimental dataset about miR-92a depletion [@helwak_mapping_2013]. It was demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions are effected depending on distance between ceRNAs or topological features of network [@nitzan_interactions_2014]. More recently, an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets was developed [@markus_list_sponge_2017] which was effective in analyzing genes through miRNAs. Based on the observations that a miRNA can exhibit strong functionality to a target but may not against an other, the authors have concluded that existing miRNA based approach may not be suitable for understanding regulations of ceRNA interactions. Additionally, miRmapper package [@da2018mirmapper] utilizes an adjacency matrix to associate miRNAs using differentially expressed genes and identifies significant miRNAs in network.

>>>>

---
title: "Network based multifactorial modelling microRNA-target interaction."
author:
- Selcen Ari^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
- Alper Yilmaz^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
  html_document: default
link-citations: yes
cls: NAR.cls
template: NAR-sample.tex
bibliography: ref.bib
---

@@

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs. We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *SERPINE2*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package [ceRNAnetsim](https://github.com/selcenari/ceRNAnetsim).

@@

### Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (\autoref{fig1}a) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (\autoref{fig1}b). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (\autoref{fig1}d), consequently have potential to affect whole network due to "ripple effect".

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). c) Distribution of miRNA1 (M1) changes. d) The change at expression of common target affects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.](Fig1.jpg){#fig1}

@@

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficieny of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network: https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. We have determined the most effective nodes in network and fold changes of genes as compared tumor and normal tissue samples of the same patient.As a result of comparison of these two datasets, we selected a gene , SERPINE2, as trigger for simulation of model (see Supplementary Materials and Methods).

@@

## Results and Conclusions
### Network with miRNA:ceRNA expression level
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (\autoref{fig1}d).

@@

[#TODO]: <> "figure1, 2 ve 3 uzerinden farkli dagilim ve degradasyon açıklaması buraya. G2 uzerinden olmasa iyi olur" 
[FIXED by selcen]
When the model run based on expression values of miRNAs and ceRNAs (genes), proportional distribution of miRNA on targets is determined in accordance with \eqref{eq:1, eq:2}. In that case, the system is observed as shown in \autoref{fig1}a at steady-state. However, if interaction factors are taken into account in the approach, specified interaction factors (i.e, binding factors such as seed type and energy) affect distribution of miRNA on targets as calculated in \eqref{eq:3} (shown in \autoref{fig2}c). After distributed miRNA efficiency was re-calculated with degradation factor(s) (i.e region effect) according to \eqref{eq:4}, steady-state condition of network is obtained (shown in \autoref{fig2}d and \autoref{fig3}a). For instance, proportional distribution of G1:M1 interaction in \autoref{fig1}a is calculated lower than the same pair (G1:M1) in the other approach that includes interaction factors (\autoref{fig2}c). But, a decrease is observed in repression efficiency of M1 miRNA on G1 in interaction factor based network approach (\autoref{fig2}d and \autoref{fig3}a).

@@

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary Materials and Methods**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

@@

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state **(see details in Supplementary Materials and Methods)**. Simulations show that change in expression level of single gene has potential to affect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

[#TODO] We have observed that the results of simulation with SERPINE2 gene, a serine protease inhibitor, one of the overexpressed genes of breast cancer [@yang2018expression; @candia2006protease] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach. Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

[#TODO]: <> "run perturbation screening on tcga set, find a gene then couple sentences about that gene"
[fixed by selcen]

@@

Network based approaches for analyzing miRNA:target interactions have been developed in earlier studies. An initial  attempt demonstrated the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs [@figliuzzi_micrornas_2013]. Next, a network based kinetic model integrating miRNA and target rates of transcription, degradation, binding and unbinding was developed [@nitzan_interactions_2014] using high throughput experimental dataset about miR-92a depletion [@helwak_mapping_2013]. It was demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions are affected depending on distance between ceRNAs or topological features of network [@nitzan_interactions_2014]. More recently, an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets was developed [@markus_list_sponge_2017] which was effective in analyzing genes through miRNAs. Based on the observations that a miRNA can exhibit strong functionality to a target but may not against an other, the authors have concluded that existing miRNA based approach may not be suitable for understanding regulations of ceRNA interactions. Additionally, miRmapper package [@da2018mirmapper] utilizes an adjacency matrix to associate miRNAs using differentially expressed genes and identifies significant miRNAs in network.
